---
title: "Forecast Analysis"
format:
  html:
    embed-resources: true
editor: visual
---

## Problem set

You will be using your model_id, climatology and persistence in the analysis below. You will be using site_ids BARC and SUGG (since they have year-around data)

Resources:

<https://frec-5174.github.io/eco4cast-in-R-book/visualizing.html>

<https://frec-5174.github.io/eco4cast-in-R-book/process-model-forecast-evaluation.html>

```{r}
library(tidyverse)
library(lubridate)
library(arrow)
```

### Question 1

Plot climatology, persistence, and your model for a single forecast day (one reference_datetime) on the same plot. Use geom_ribbon to plot the uncertainty, geom_line to plot the median, and geom_point to plot the observations.

```{r}
# Use code below

JM <- arrow::open_dataset("s3://anonymous@bio230014-bucket01/challenges/scores/parquet/project_id=neon4cast/duration=P1D/variable=temperature/model_id=TSLM_seasonal_JM?endpoint_override=sdsc.osn.xsede.org") |> 
  dplyr::collect() |> 
  mutate(model_id = "TSLM_seasonal_JM")

climatology <- arrow::open_dataset("s3://anonymous@bio230014-bucket01/challenges/scores/parquet/project_id=neon4cast/duration=P1D/variable=temperature/model_id=climatology?endpoint_override=sdsc.osn.xsede.org") |> 
  dplyr::collect() |> 
  mutate(model_id = "climatology")

persistenceRW <- arrow::open_dataset("s3://anonymous@bio230014-bucket01/challenges/scores/parquet/project_id=neon4cast/duration=P1D/variable=temperature/model_id=persistenceRW?endpoint_override=sdsc.osn.xsede.org") |> 
  dplyr::collect() |> 
  mutate(model_id = "persistenceRW")

alldata <- bind_rows(JM, climatology, persistenceRW)
alldata$reference_datetime <- ymd(alldata$reference_datetime)
alldata$date <- ymd(alldata$date)

alldata |> summarise(min = min(reference_datetime),
                     max = max(reference_datetime),
                     .by = model_id)

filter(alldata, reference_datetime ==  "2024-03-04", site_id == "BARC") |> 
ggplot() +
  geom_line(aes(x = date, y = median, color = model_id))+
  geom_ribbon(aes(x = date, ymin = quantile02.5, ymax = quantile97.5, color = model_id, fill = model_id), alpha = 0.2)+
  geom_point(aes(x = date, y = observation))+
  labs(title = "Model predictions for BARC")

```

### Question 2

Based on visual inspection of your plot, how do the median of models differ in how they represent the observations.
  
The climatology median is very close to the actual temperatures observed. At some points my model does better, but at many times it is worse. The persistence model is very consistently slightly below the actual temperature.   

### Question 3

Based on visual inspection of your plot, how does the uncertainty of each model forecasts differ in capacity to represent the observations.
  
The persistence model has the most uncertainty, and it gets significantly wider as the horizon increases. This successfully captures all of the observations. The climatology does not have uncertainty that propagates, so the confidence interval doesn't grow as the horizon increases. Despite this, it still captures the observations successfully. My model does not have enough uncertainty, which often results in the observations falling outside of my confidence interval, so it does a poor job of representing the observations.  

### Question 4

Calculate the mean CRPS for the three models(averaged across all horizons, sites, and reference_datetimes). Which model has the lower score?

```{r}


```

### Question 5

Plot the mean CRPS vs horizon for all three models. How does the performance change as you forecast further in the future?

```{r}


```

### Question 6

Plot the mean CRPS separately for each site_id for all three models. How does performance differ between sites?

```{r}


```

### Question 7

Which forecasting best practices are addressed with your forecasts and the analysis above? See <https://frec-5174.github.io/eco4cast-in-R-book/best-practices.html>.
